---
title: "Homework #03: Community-Based Intervention and  Child Social Development"
subtitle: "due Friday, November 11 at 4:00pm"
output: 
  tufte::tufte_html:
    css: "sta199-labs.css"
    tufte_variant: "envisioned"
    highlight: pygments
    toc: true
    toc_depth: 1
link-citations: yes
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Miller et al (2021) report the results of a randomized, controlled trial designed to evaluate effectiveness of a large-scale community-based group parenting intervention (CASITA) in the Carabayllo distrit of Lima, Peru in 2017-2018. The intervention, designed to increase the caregiver's ability to provide developmentally-stimulating interactions with their young child, included parent coaching and social support. The Extended Ages and Stages Questionnaire (EASQ) was used to measure child development at baseline and at the end of the study. Eligible children were between 6-20 months and either at risk of developmental delay or with developmental delay already identified. The enrolled children were randomized with equal probability into the CASITA intervention group or a control group (note the control families all were offered the intervention at the end of the study period). 

The data analysis plan specified that the primary response would be EASQ age-standardized z-scores at follow-up (`total_z_vs_wb2`) and that the data would be analyzed using linear regression analysis on the intent-to-treat population, with covariates in the model including the following:

- treatment group `study_arm1` (0=control, 1=intervention)

- maternal education `mom_educ` (0=primary, 1=secondary, or 2=post-secondary) 

- gestational age of the child at birth `gest_age_cat` (0=40+weeks, 1=38-39 weeks, 2=36-37 weeks)

- child sex `Sex` (0=missing, 1=male, 2=female) 

- baseline age-standardized EASQ z-score `total_z_vs_wb1`

While an appropriate t-test would be a valid way of analyzing the results of this randomized study, it was not prespecified as the primary analysis here because the researchers wanted to control for known variability in development as a function of child sex, gestational age at birth, and other factors, which should lead to (appropriately) more tighter confidence intervals.


In this homework, we do not consider the full EASQ but instead consider the social development subscale, contained in the variables `persoc_z_vwb1` at baseline and `persoc_z_wb2` at follow-up.

```{r readdata, message=FALSE, warning=FALSE}
library(haven) #to read the Stata .dta file
library(tidyverse)
library(tidymodels)
cdev <- read_dta("cdev.dta")
```


1. Before doing any analysis, you need to restrict the data to include only children with `study_a_participant` value=3, as some data may be provided for children who did not officially enroll in the study. Additionally, you will need to filter out observations where there are missing labels for treatment group. Finally, to prepare for fitting a linear regression model later, each of the categorical variables with many levels will need to be converted into dummy variables. Each of the dummy variables will correspond to one level of a categorical variable. (Refer to the end of Week 9 Lecture 1 for a refresher). Example code to encode `mom_educ` is provided below. Note that we only make dummy variables for two out of the three possible levels (secondary & post-secondary). Therefore, the reference group when interpreting regression output later will be the excluded one (in this case primary education). Please do the same in making dummy variables for `gest_age_cat` and `Sex`.  Be sure to display your code in your submission.

```{r chunklabelyoulike, eval=FALSE}
# example code 
cdev %>%
  # this corresponds to the two levels of mom_educ
  mutate(edsecondary=ifelse(mom_educ==1,"yes","no")) %>%
  mutate(edpostsecondary=ifelse(mom_educ==2,"yes","no")) 
  
 
```


2. The social development scores were converted into z-scores using age-specific quantiles obtained from a Peruvian population. (Age-specific z-scores are standardized to have mean 0 and variance 1 separately for each age group, so that you can directly compare children of different ages -- an "average" child at age 8 months will have z-score of 0, and an "average" child at age 1 year will also have a z-score of 0, even though the raw scores may differ as children develop). Provide a histogram of baseline social development z-scores. What proportion of children scored above the age-adjusted mean on the social development scale at baseline? Do these scores appear to represent children who have developmental delay or who are at risk of delay?  

3. Create a scatterplot that has the baseline social development z-scores on the x-axis and the follow-up social development scores on the y-axis. Make sure your plot makes it easy to distinguish treatment (CASITA) versus control participants in the study.  Does the plot help you determine whether CASITA is effective?  Explain why or why not.



4. Create a 95% confidence interval for the change in z-score from baseline to follow-up in the control group, as well as a 95% confidence interval for the change in z-score from baseline to follow-up in the CASITA treatment group. Conduct a t-test for (a) the null hypothesis that age-adjusted z-scores social development were unchanged from baseline to follow-up in the control group, and (b) the null hypothesis that age-adjusted z-scores for social development were unchanged from baseline to follow-up in the CASITA group. Clearly specify what type of t-test you used, providing a brief rationale. Briefly describe the substance of your findings.


5. What kind of t-test is most appropriate for determining whether the treatment is effective, using a direct comparison of changes in CASITA versus control? Provide a justification for your preferred t-test, conduct it, and interpret the results, including a single 95% confidence interval corresponding to the test you conducted.

6. Fit the linear regression model specified in the data analysis plan, except that you should use social development scores rather than full EASQ scores.  Provide a plot of residuals vs predicted follow-up social development scores and note whether you have any concerns based on the plot.  


7. Is CASITA effective at increasing social development?  Provide evidence (e.g., a test result or confidence interval) based on the regression model to support your claim, and if CASITA is effective, provide an estimated benefit in terms of the difference in z-score, holding the other factors in the model constant.


8. Are any other factors in the model "significantly" related to follow-up social development at an $\alpha=0.05$ significance level?  For each "significant" factor, describe how that factor is related to follow-up social development using estimates from your model.


9. Does the model as a whole explain most of the variability in the social development scores at follow-up? Provide statistical evidence to support your answer.

10. Researchers would like to check whether the CASITA intervention is equally effective for participants regardless of baseline social development score. Add an interaction term to your model that allows you to evaluate this question, perform a test of the null hypothesis that the CASITA intervention is equally effective for participants regardless of baseline social development score, and interpret the results of your test.


### Grading:

| Component | Points |
|:----------|:-------|
| Ex 1      | 2     |
| Ex 2      | 3     |
| Ex 3      | 5     |
| Ex 4      | 5     |
| Ex 5      | 5     |
| Ex 6      | 5     |
| Ex 7      | 5     |
| Ex 8      | 5     |
| Ex 9      | 5     |
| Ex 10     | 5     |
| Workflow & formatting   | 5      |

**Grading notes**:

-   The "Workflow & formatting" grade is to assess the reproducible workflow. This includes having at least 3 informative commit messages, labeling the code chunks, updating the name and date in the YAML, and in general having a neat report.
